---
title: "Cleaning Variable Names with Janitor"
author: "C.M. Gienger"
date: "`r Sys.Date()`"
output: html_notebook
---

```{r setup, include=FALSE}
#Its common to have a setup area to help keep track of required packages and formatting options
library(tidyverse) #importing, tidying, plotting data
library(knitr) #making tables
library(janitor) #cleaning variable names
library(readxl) #importing data from excel spreadsheets.
```
A common best-practice is to make sure variable names have consistent formatting. This includes having
a common case of text, common delimiters, and removal of punctuation and other nusiance charaters.

Example data are from [Ord and Klomp 2014](https://link.springer.com/content/pdf/10.1007/s00442-014-2921-y.pdf)  

```{r import raw data}
data_lizards_raw <- read_excel("data/data_morph_lizards.xlsx")
```

It already detects that we have duplicate column names in the data.

```{r clean the data}
data_lizards_clean <- read_excel("data/data_morph_lizards.xlsx") %>% 
  clean_names() %>% #makes variable have consistent formatting and pretty
  remove_empty(which = c("rows","cols")) # remove both empty columns and rows; happens a lot when importing Excel data
```

What does this do?

- Returns names with only lowercase letters, with _ as a separator
- Handles special characters and spaces
- Appends numbers to duplicated column names
- Converts “%” to “percent” to retain meaning
- Removes empty columns and rows that often persist in Excel sheets

Lets compare the raw versus the cleaned data.

```{r echo=FALSE}
combined <- data.frame(raw = colnames(data_lizards_raw), cleaned = colnames(data_lizards_clean))
kable(combined, caption = 'Table 1. A knitr kable displaying the raw vs. cleaned data.')
```
Previous warning message told us the SVL variable is listed twice. Lets remove it and fix SVL label in column 5.

```{r}
data_lizards_clean <- select(data_lizards_clean, -svl_mm_14) %>% #drop column svl_mm_14
  rename(svl_mm = svl_mm_5)
```

Lets give the data a quick summary look.

```{r}
tabyl(data_lizards_clean, species)
```
OK, looks pretty good. One last thing is to check whether there are any duplicate records that should
be excluded.

```{r}
data_lizards_clean %>% janitor::get_dupes(id)
```

Can exclude the two duplicates.

```{r}
data_lizards_clean <- dplyr::distinct(data_lizards_clean) #takes only the unique observations.
```
Notice that the number of observations went from 251 (with duplicates) to 249 (no duplicates).

```{r}
head(data_lizards_clean) #first six obs. of cleaned data set; no wonky names, no dupes, consistent formatting
```

